---
title: "Crops01"
author: "Justin Maynard, Ellie Shang & Pablo Salazar"
date: "2025-04-14"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
library(readxl)
library(lubridate)
library(ggplot2)
library(forecast)
library(cowplot)
library(tseries)
library(Kendall)
library(agridat)
library(outliers)
library(smooth)
library(kableExtra)
#library(tidyverse)
```
```{r}
# Set the path to your desktop (Windows example)
library(here)
here()
file_path <- paste0(here(),"/data/","Wheat Data-All Years.xlsx")



# Read the Excel file
data_rye <- read_excel(file_path, sheet = "Table02", skip = 1)
data_wheat <- read_excel(file_path, sheet = "Table04", skip = 1)

```

```{r}
wheat_p <- data_wheat$`U.S. production (million bushels)`  # Replace with the exact column name if different

# Create a time series starting from 1866 with yearly data
start_year_w <- 1960
wheat_ts <- ts(wheat_p, start = start_year_w, frequency = 1)


# Optionally, plot the time series
plot(wheat_ts, main = "U.S. production (million bushels)", ylab = "U.S. production (million bushels)", xlab = "Year")
```


```{r}
yield_data <- data_rye$`Yield (bushels per acre)`  # Replace with the exact column name if different

# Create a time series starting from 1866 with yearly data
start_year <- 1866
rye_ts <- ts(yield_data, start = start_year, frequency = 1)

# View the time series
print(rye_ts)

# Optionally, plot the time series
plot(rye_ts, main = "Rye Yield (Bushels per Acre)", ylab = "Yield", xlab = "Year")
```


```{r}
rye_ACF <- Acf(rye_ts, lag=40, plot=FALSE)
rye_PACF <- Pacf(rye_ts, lag=40, plot=FALSE)

autoplot(rye_ACF) + theme_classic()
autoplot(rye_PACF) + theme_classic()
```


```{r}
rye_ts

```
```{r}
rye_tsc <- na.omit(rye_ts)
length(rye_tsc)
```

```{r}
nobs <- length(rye_tsc)
t <- c(1:nobs)

linear_trend_model_rye <- lm(rye_tsc ~ t) 
summary(linear_trend_model_rye)

#saving regression coefficients
beta0_rye <- as.numeric(linear_trend_model_rye$coefficients[1])
beta1_rye <- as.numeric(linear_trend_model_rye$coefficients[2])
```

> To improve the accuracy and reliability of our time series forecasting for U.S. rye yields (measured in bushels per acre), we deliberately limited the dataset to include only data from 1920 onward. This decision was made to minimize potential skewness introduced by earlier, less consistent historical data, and to better reflect the agricultural practices, technologies, and climate conditions relevant to the modern era.


```{r}
# Original cleaned time series
start_year <- 1866
new_start <- 1920
new_end <- 2024

# Calculate the index position of the new year
start_index <- new_start - start_year + 1

# Create new time series from new year onward
rye_ts_ny <- window(rye_tsc, start = new_start, end = new_end)
```

```{r}
rye_ts_ny
```


```{r}
nobs <- length(rye_ts_ny)
t <- c(1:nobs)

linear_trend_model_rye_ny <- lm(rye_ts_ny ~ t) 
summary(linear_trend_model_rye_ny)

#saving regression coefficients
beta0_rye <- as.numeric(linear_trend_model_rye_ny$coefficients[1])
beta1_rye <- as.numeric(linear_trend_model_rye_ny$coefficients[2])
```



State space exponential smoothing
```{r}
#Plotting Data

ts <- autoplot(rye_ts, plot = FALSE) + theme_classic()

pacf <- autoplot(Pacf(rye_ts, lag = 40, plot=FALSE),  
                  main = "PACF") + theme_classic()

acf <- autoplot(Acf(rye_ts, lag = 40, plot=FALSE),  
                  main = "ACF") + theme_classic()

plot_grid(ts, pacf, acf,
          nrow = 3
)


```

```{r}
#Stationary Test

print((adf.test(rye_ts_ny,alternative="stationary")))
#P value less than .05 so we fail to reject H0, data has no stochastic trend
```
Forecasting Simple Techniques
```{r}
naive_rye <- naive(rye_ts_ny, h = 5)
plot(naive_rye)

```

```{r}
#simple moving average
sma_rye <- sma(y = rye_ts_ny, h = 5, holdout = FALSE, silent = FALSE)
plot(sma_rye)

autoplot(rye_ts_ny) + 
  autolayer(sma_rye$forecast)
```

```{r}
#simple exponential smoothing
ses_rye <- ses(y = rye_ts_ny, h = 5, holdout = FALSE, silent = FALSE)

plot(rye_ts_ny, col = 'black')
lines(ses_rye$fitted, col = 'blue')
```

```{r}
#auto arima
rye_ARIMA <- auto.arima(rye_ts_ny,seasonal = FALSE, lambda = 0)
rye_arima_forecast <-forecast(rye_ARIMA)

autoplot(rye_arima_forecast, ylab = "Yield (bushels per acre)")

arima_scores <- accuracy(rye_arima_forecast)

```



BATS

```{r}
TBATS_fit <- tbats(rye_ts_ny)
TBATS_forecast <- forecast(TBATS_fit)

autoplot(TBATS_forecast) + 
  autolayer((rye_ts_ny))

TBATS_scores <- accuracy(TBATS_forecast)

```

Neural Network

```{r}
NN_fit <- nnetar(rye_ts_ny,
                 P = 2,
                 p = 2)
                
NN_forecast <- forecast(NN_fit) 

autoplot(NN_forecast) + 
  autolayer(rye_ts_ny)

NN_scores <- accuracy(NN_forecast)

```

State space exponential smoothing
```{r}
SSES_rye <- es(rye_ts_ny, model = "ZZZ", h <- 10, holdout = FALSE)
plot(SSES_rye)
checkresiduals(SSES_rye)
SSES_rye$model

SS_fit <- StructTS(rye_ts_ny,
                   type = "trend")
checkresiduals(SS_fit)

SS_forecast <- forecast(SS_fit, h = 10)

autoplot(SS_forecast) +
  autolayer(rye_ts_ny)

SS_scores <- accuracy(SS_forecast)

```

```{r}
# Combine all scores into one data frame
all_scores <- rbind(
  cbind(Model = "ARIMA", arima_scores),
  cbind(Model = "TBATS", TBATS_scores),
  cbind(Model = "NNETAR", NN_scores),
  cbind(Model = "StructTS", SS_scores)
)

# Convert to data frame for kable
all_scores <- as.data.frame(all_scores)

# Ensure numeric columns are correctly typed (in case they are characters from rbind)
numeric_cols <- setdiff(colnames(all_scores), "Model")
all_scores[numeric_cols] <- lapply(all_scores[numeric_cols], as.numeric)

# Find index of the row with the lowest RMSE
lowest_rmse_index <- which.min(all_scores[["RMSE"]])

# Create a styled table with the model that has the lowest RMSE highlighted
kbl(all_scores, 
    caption = "Forecast Accuracy Across Models",
    digits = array(5, ncol(all_scores))) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  row_spec(lowest_rmse_index, bold = TRUE, background = "#DFF0D8") %>%
  kable_styling(latex_options = "striped")
```


>> Looking at the accuracy results for the four different time series models—ARIMA, TBATS, NNETAR, and StructTS— that were evaluated to forecast rye yields in the U.S. over the next 10 years. We can see that NNETAR stand out. Each model was assessed using standard accuracy metrics, including Mean Error (ME), Root Mean Squared Error (RMSE), Mean Absolute Error (MAE), and Mean Absolute Scaled Error (MASE), among others.

>> The NNETAR model has the lowest RMSE (1.91) and MAE (1.55), indicating superior forecasting accuracy compared to the other models. It also posted the lowest MASE (0.82), reinforcing its relative strength in capturing patterns in the training data. Although its Mean Percentage Error (MPE) was slightly negative (-1.25), this implies a modest tendency to underpredict.

>> The ARIMA model performed reasonably well, with a slightly higher RMSE (1.99) and MAE (1.66), but it maintained a positive MPE (1.02) and low ACF1 (0.04), suggesting minimal autocorrelation in residuals—an indicator of good model fit.

>> StructTS and TBATS both underperformed relative to NNETAR and ARIMA, with RMSE values exceeding 2.0 and higher MAE and MAPE scores. While TBATS had the highest ME and RMSE, its negative ACF1 value (-0.089) may point to some overfitting or residual autocorrelation that undercuts forecast reliability.

